---
title: "2.R_Visualization"
author: "Kim Wong"
date: "09/16/2021"
output:
  pdf_document: default
  html_notebook: default
---
##
## Material taken from
## Graham J. Williams. 2017. The Essentials of Data Science: Knowledge Discovery Using R (1st ed.). Chapman & Hall/CRC.
##

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load required packages
```{r}
library(dplyr)        # glimpse().
library(ggplot2)      # Visualise data.
library(magrittr)     # Data pipelines: %>% %<>% %T>% equals().
library(randomForest) # na.roughfix() for missing data.
library(rattle)       # normVarNames().
library(rattle.data)  # weatherAUS.
library(scales)       # commas(), percent().
library(stringr)      # str_replace_all().
library(stringi)      # String concat operator %s+%.
```

## Edgar Anderson's Iris Data
```{r}
iris
```


## A Simple scatterplot 
```{r}
iris %>%
  ggplot(aes(x=Sepal.Length, y=Sepal.Width, size=4)) +
  geom_point()
```

## A simple line plot with data points
```{r}
iris %>%
  ggplot(aes(x=Sepal.Length, y=Sepal.Width)) +
  geom_point() +
  geom_line()
```


## A Simple scatterplot but color coding of points based on species variable; also encode size
```{r}
iris %>%
  ggplot(aes(x=Sepal.Length, y=Sepal.Width, color=Species)) +
  geom_point(aes(size=Petal.Length))
```

## A Simple scatterplot but shape coding of points based on species variable
```{r}
iris %>%
  ggplot(aes(x=Sepal.Length, y=Sepal.Width, shape=Species, size=Petal.Width)) +
  geom_point()
```

## dimensions of data frame

```{r}
iris %>%
  ggplot(aes(x=Sepal.Length, y=Sepal.Width)) +
  geom_point() +
  #stat_smooth(method="lm", formula=y~x+I(x^2))
  #stat_smooth(method="lm", formula=y~poly(x,2))
  #stat_smooth(method="loess", formula=y~x)
  stat_smooth(method="gam", formula=y~s(x), se=T)
```



## Let's load the data which we output from the data cleaning notebook
```{r}
fpath <- getwd() %>% print()
```
## generate timestamp
```{r}
#dsdate  <- "_" %s+% format(Sys.Date(), "%Y%m%d") %T>% print()
dsdate <- "_20201104"
```
## specify filename for dataset
```{r}
dsname="cleaned_weatherAUS"
dsrdata <- 
  file.path(fpath, dsname %s+% dsdate %s+% ".RData") %T>% 
  file.path(fpath, dsname %s+% dsdate %s+% ".RData") %T>% 
  print()
```

## Load R objects from file and list them
```{r}
load(dsrdata) %>% print()
```

## Take a peak at the data
```{r}
glimpse(ds)
```
## Count the number of missing values.
```{r}
ds[vars] %>% is.na() %>% sum() %>% comcat()
```

## Impute missing values.
```{r}
ds[vars] %<>% na.roughfix()
```

## Confirm that no missing values remain
```{r}
ds[vars] %>% is.na() %>% sum() %>% comcat()
```

## Take a peak at the data again
```{r}
glimpse(ds)
```

## Randomly pull out a sample of 1000 observations from data set
```{r}
ds %>% nrow() %>% sample(1000) -> sobs
```
## Do a scatterplot of the min and max temperatures and encode if it will rain tomorrow in the color
```{r}
ds %>% extract(sobs,) %>% 
  ggplot(aes(x=min_temp, y=max_temp, color=rain_tomorrow)) +
  geom_point()
```

## Standard bar chart.
```{r}
ds %>% ggplot(aes(x=wind_dir_3pm)) + geom_bar()
```

## ## Standard bar chart with binary outcome encoded in the color.
```{r}
ds %>% ggplot(aes(x=wind_dir_3pm, fill=rain_tomorrow)) + 
  geom_bar()
```

## Define two shades of blue using ColorBrewer, http://colorbrewer2.org.

```{r}
library(RColorBrewer)
blues2 <- brewer.pal(4,"Paired")[1:2] %T>% print()
```
## Output the number of weather stations from data set
```{r}
ds$location %>%
  unique() %>%
  length() %>%
  print() ->
  num_locations
```

## Generate a bar chart with more complex annotation
```{r}
ds %>%
  ggplot(aes(x=wind_dir_3pm, fill=rain_tomorrow)) +
  geom_bar() +
  scale_fill_manual(values = blues2,
                    labels = c("No Rain", "Rain")) +
  scale_y_continuous(labels = comma) +
  theme(legend.position = c(0.95, 0.85),
        legend.title      = element_text(color="grey40"),
        legend.text       = element_text(color="grey40"),
        legend.background = element_rect(fill="transparent")) +
  labs(title    = "Rain Expectedd by Wind Direction at 3PM",
       subtitle = "Observations from " %s+%
                   num_locations %s+% "weather stations",
       caption  = "Source: Australian Bureau of Meteorology",
       x        = "Wind Direction 3PM",
       y        = "Number of Days",
       fill     = "Tomorrow")
```
## Bar chart of mean temperature at 3PM across all 49 weather stations
```{r}
ds %>%
  ggplot(aes(x=location, y=temp_3pm, fill=location)) +
  geom_bar(stat="summary", fun.y="mean") +
  theme(legend.position="none")
```
## Since the above plot has all labeled weather stations cluttered together, let's rotate the labels by 90 degrees.
```{r}
ds %>%
  ggplot(aes(x=location, y=temp_3pm, fill=location)) +
  geom_bar(stat="summary", fun.y="mean") +
  theme(legend.position="none") +
  theme(axis.text.x = element_text(angle=90))
```
## Instead of rotating the labels by 90 degrees, we can flip the x-y coordinates.
```{r}
ds %>%
  ggplot(aes(x=location, y=temp_3pm, fill=location)) +
  geom_bar(stat="summary", fun.y="mean") +
  theme(legend.position="none") +
  #theme(axis.text.x = element_text(angle=90)) +
  coord_flip()
```

## We want to reorder the location in alphabetical order instead of the reverse
```{r}
ds$location %>%
  levels() %>%
  rev() ->
  loc
```

## Use mutate within dplyr to temporarily place the location in alphabetical order
```{r}
ds %>%
  mutate(location=factor(location,levels=loc)) %>%
  ggplot(aes(x=location, y=temp_3pm, fill=location)) +
  geom_bar(stat="summary", fun.y="mean") +
  theme(legend.position="none") +
  #theme(axis.text.x = element_text(angle=90)) +
  coord_flip()
```

## Output box plot of maximum temperature as a parameter of year
```{r}
ds %>%
  ggplot(aes(year, max_temp, fill=year)) +
  geom_boxplot(notch = T) + 
  theme(legend.position="none")
```
## Output violin plot of maximum temperature as a parameter of year
```{r}
ds %>%
  ggplot(aes(year, max_temp, fill=year)) +
  geom_violin() + 
  theme(legend.position="none")
```
## Overlay box plot and violin plot of maximum temperature as a parameter of year
```{r}
ds %>%
  ggplot(aes(year, max_temp, fill=year)) +
  geom_violin() + 
  geom_boxplot(width=0.5, position = position_dodge(width=0)) +
  theme(legend.position="none")
```
## Use facet_wrap to split the plots across locations
```{r}
ds %>%
  ggplot(aes(x=year, y=max_temp, fill=year)) +
  geom_violin() + 
  geom_boxplot(width=0.5, position=position_dodge(width=0)) +
  theme(legend.position="none") +
  theme(axis.text.x=element_text(angle=45, hjust=1)) +
  facet_wrap(~location, ncol=5)
```
## Create a helper funtion so as to visualize clusters of location
```{r}
myplot <- function(ds, n)
{
ds %>%
  filter(cluster==n) %>%
  ggplot(aes(x=year, y=max_temp, fill=year)) +
  geom_violin() + 
  geom_boxplot(width=0.5, position=position_dodge(width=0)) +
  theme(legend.position="none") +
  theme(axis.text.x=element_text(angle=45, hjust=1)) +
  facet_wrap(~location, ncol=5)
}
```


```{r}
myplot(ds,"area3")
```

## convert variables from character to factor class
```{r}
lblr <- function(x) 
{
  x %>%
    str_replace_all("n", "North ") %>%
    str_replace_all("s", "South ") %>%
    str_replace_all("e", "East ") %>%
    str_replace_all("w", "West ") %>%
    str_replace(" $", "")
}
```

## convert wind direction to factor
```{r}
ds %>%
  sample_n(10000) %>%
  ggplot(aes(x=min_temp, y=max_temp, colour=rain_tomorrow)) +
  geom_point(shape=".") +
  geom_smooth(method="gam", formula=y~s(x, bs="cs")) +
  facet_wrap(~wind_dir_3pm, labeller=labeller(wind_dir_3pm=lblr)) +
  labs(x      = "Min Temp", 
       y      = "Max Temp", 
       colour = "Rain Tomorrow")
```
























